Systemix Article: Combined MAG-SMAG Profiling

### 2024-04-29 R/paper.Rmd
updated README (section 3), figures and Rdata

---
title: "Systemix Profiling Article: Combined MAG-SMAG Analysis"
author: "Moises A. Rojas"
date: "2023-12-01"
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_knit$set(root.dir = '/mnt/beegfs/home/mrojas/DiGenomaLab/Systemix/analysis/mags-all/analysis/')

suppressMessages(library(ape)) #tree basics
suppressMessages(library(treeio))
suppressMessages(library(ggtree))
suppressMessages(library(ggplot2))
suppressMessages(library(ggtreeExtra)) #geom_fruit
suppressMessages(library(ggnewscale)) #scales in ggplot2
suppressMessages(library(ggstar)) #geom_star for ggplot2
suppressMessages(library(phytools)) #MRCA
suppressMessages(library(tidyverse)) #strings, dataframes processing
suppressMessages(library(magrittr)) #pipes %>%
suppressMessages(library(tidytree)) #trees as dataframes
suppressMessages(library(RColorBrewer)) #colors palette
suppressWarnings(library(colorspace))
suppressWarnings(library(patchwork)) #side-by-side plotting
```

@@

```{r magsdata}
library(openxlsx)

# Read Systemix MAGs taxonomy and QC files
gtdb_taxonomy <- read.csv("1-taxonomy/gtdbtk_classification.tsv", sep = "\t", stringsAsFactors = FALSE)
mags_qc <- read.csv("2-statistics/checkm_parsed_bins.tsv", sep = "\t")

@@

# Parse and reorder columns
mag_samples <- dplyr::select(mag_samples, -c("taxonomy", "gc", "contigs"))
mag_samples <- mag_samples %>%
  mutate_if(is.character, function(x) gsub("d__|p__|c__|o__|f__|g__|s__", "", x))
mag_samples <- mag_samples %>% separate(bin, "\\.", into = c("mag_name", "mag_id"))
mag_samples <- mag_samples[, c(1, 2, 10, 11, 12, 13, 3, 4, 5, 6, 7, 8, 9)]

# Define final attributes
mag_samples['ecosystem'] <- "Biomining"
mag_samples['continent'] <- "South America"

# Load SMAG metadata (tables 1 and 2)
smag_md1 <- read.xlsx("/mnt/beegfs/home/mrojas/DiGenomaLab/Systemix/SMAG/Extended_Data_Table_1.xlsx")
smag_md2 <- read.xlsx("/mnt/beegfs/home/mrojas/DiGenomaLab/Systemix/SMAG/Extended_Data_Table_2.xlsx")

# Save attributes
smag_md1_sel <- smag_md1 %>% dplyr::select("mag_name", "Ecosystem", "Continent")
smag_md2_sel <- smag_md2 %>%
  dplyr::select("user_genome", "completeness", "contamination",
                "strain_heterogeneity", "N50", "classification",
                "phylum", "class", "order", "family", "genus", "species")

@@

# Rename columns
colnames(smag_data)[7] <- "kingdom"
colnames(smag_data)[14] <- "ecosystem"
colnames(smag_data)[15] <- "continent"

# Clean memory
rm(smag_md1, smag_md2, smag_md1_sel, smag_md2_sel, mags_qc)

@@

# Save parsed SMAG metadata
if (file.exists("4-mags-smag/smag_data.tsv") == FALSE) {
  write.table(smag_data, "4-mags-smag/smag_data.tsv",
              sep = "\t", quote = FALSE, row.names = FALSE)
}
```

## In-depth taxonomic analysis of MAG-SMAG

```{r magstaxa}
# Biomining provides a new phylum -> Nitrospirota_A
mag_samples_new_phy <- mag_samples[!(mag_samples$phylum %in% smag_data$phylum), ]

# Explore metadata: some bins are not assigned to ecosystems
mags_metadata %>%
  filter(is.na(ecosystem)) %>%
  group_by(mag_name) %>%
  dplyr::summarise() #n=401

# Keep only bacteria and defined ecosystems (discard archaea kingdom and non-NA eco.)
mags_metadata_bac <- mags_metadata %>%
  filter(kingdom == "Bacteria" & !is.na(ecosystem)) #n=32975

# Save to file
if (file.exists("4-mags-smag/mags_metadata_bac.tsv") == FALSE) {
  write.table(mags_metadata_bac, "4-mags-smag/mags_metadata_bac.tsv",
              sep = "\t", quote = FALSE, row.names = FALSE)
}

# Select top phyla and plot a histogram
mags_md_top_phy <- table(mags_metadata_bac$phylum) %>%
  sort(TRUE) %>%
  head(n = 30) %>%
  as.data.frame() %>%
  dplyr::rename(phy_top = Var1, freq = Freq)

plot_depth_top_phyla <- ggplot(mags_md_top_phy, aes(x = reorder(phy_top, freq), y = freq)) +
  geom_bar(stat = "identity", fill = "#459682") +
  geom_text(aes(label = freq), vjust = 0.5, hjust = -0.5, size = 3) +
  labs(x = "Phylum", y = "Number") +
  scale_y_continuous(limits = c(0, 10000)) +
  theme_classic() +
  coord_flip()
ggsave("plot_md_top_phyla.pdf", plot_depth_top_phyla, height = 7, width = 6)

## Filtering metadata

# Set colors for ecosystem
#mags_colors_eco <- vector()
#for (i in 1:length(kegg_colors_eco)) {
#  mags_colors_eco <- append(mags_colors_eco, kegg_colors_eco[[i]])
#}
#mags_colors_eco <- colorRampPalette(brewer.pal(12, "Paired"))(length(unique(mags_metadata_bac$ecosystem)))
#mags_colors_eco <- sample(mags_colors_eco)
mags_colors_eco <- c("#1F78B4", "#FF7F00", "#B2DF8A", "#E31A1C", "#A6CEE3",
                     "#6A3D9A", "#33A02C", "#CAB2D6", "#FDBF6F", "#FB9A99")

# Which bacteria from biomining sites are annotated as known?
mags_depth_spec_biom <- mags_metadata_bac %>%
  dplyr::select(mag_name, family, genus, species, ecosystem) %>%
  filter(species != "" & ecosystem == "Biomining")

#are they present in other SMAG?
mags_depth_spec <- mags_metadata_bac[mags_metadata_bac$species %in%
                                     mags_depth_spec_biom$species, ] %>%
  dplyr::select(mag_name, mag_id, phylum, family, genus, species, ecosystem, continent) %>%
  group_by(ecosystem)

plot_depth_spec <- ggplot(mags_depth_spec) +
  geom_bar(aes(x = species, fill = phylum)) +
  labs(x = "", y = "Number of known species") +
  scale_fill_manual(values = "#DD7E8B") +
  theme_classic() +
  theme(legend.title = element_blank(),
        legend.position = c(0.85, 0.05),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  facet_wrap(.~ecosystem)
ggsave("plot_md_biom_species.pdf", plot_depth_spec, height = 6, width = 8)

# Which ecosystem has more known species (# and ratio)?
mags_depth_spec_eco <- mags_metadata_bac %>%
  dplyr::filter(species != "") %>%
  group_by(species, ecosystem) %>%
  dplyr::summarise(count_sp = table(species)) %>% #.groups = "drop"
  ungroup() %>%
  group_by(ecosystem) %>%
  dplyr::summarise(count_eco = sum(count_sp)) %>%
  dplyr::arrange(desc(count_eco)) #max 1821

@@

# What about unknown/known species-level genome bins (SGBs) at phylum level?
mags_depth_sgbs <- mags_metadata_bac %>%
  group_by(phylum, species) %>%
  dplyr::summarise(n_spe = n()) %>%
  ungroup() %>%
  group_by(phylum) %>%
  dplyr::mutate(t_phy = sum(n_spe)) %>% ###
  dplyr::filter(species == "") %>%
  dplyr::mutate(diff_k = t_phy - n_spe) %>%
  dplyr::mutate(perc_u = n_spe * 100 / t_phy) %>%
  dplyr::arrange(desc(t_phy)) %>%
  pivot_longer(cols = c(n_spe, diff_k), names_to = "cond", values_to = "tot")

@@

plot_depth_sgbs_p <- ggplot(mags_depth_sgbs_p[1:60, ],
                            aes(x = reorder(phylum, t_phy), y = tot_p, fill = cond_p)) +
  geom_col(position = position_stack(reverse = TRUE)) +
  scale_fill_manual(name = "", labels = c("kSGB", "uSGB"),
                    values = c("#DD7E8B", "#459682")) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(x = "", y = "% of species") +
  theme_classic() +
  theme(legend.position = "none",
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.line.y = element_blank()) +
  coord_flip()
#ggsave("plot_md_SGBs_phylum_p.pdf", plot_depth_sgbs_p, height = 7, width = 5)

@@

# And unknown/known SGBs for biomining?
mags_depth_sgbs_biom <- mags_metadata_bac %>%
  dplyr::filter(ecosystem == "Biomining") %>%
  group_by(phylum, species) %>%
  dplyr::summarise(n_spe = n()) %>%
  ungroup() %>%
  group_by(phylum) %>%
  dplyr::mutate(t_phy = sum(n_spe)) %>%
  dplyr::filter(species == "") %>%
  dplyr::mutate(diff_k = t_phy - n_spe) %>%
  dplyr::mutate(perc_u = n_spe * 100 / t_phy) %>%
  dplyr::arrange(desc(t_phy)) %>%
  pivot_longer(cols = c(n_spe, diff_k), names_to = "cond", values_to = "tot")

@@

# What is the distribution of uSGBs by ecosystem?
mags_depth_usgbs_eco <- mags_metadata_bac %>%
  dplyr::filter(species == "") %>%
  group_by(phylum, ecosystem) %>%
  dplyr::summarise(num = table(ecosystem)) %>%
  dplyr::mutate(phylum = replace(phylum, num < 5, "Other")) #group lows as "other"

plot_depth_usgbs <- ggplot(mags_depth_usgbs_eco,
                           aes(x = reorder(phylum, num), y = num, fill = ecosystem)) +
  geom_col() +
  labs(x = "Phylum", y = "Number of uSGBs") +
  scale_y_continuous(n.breaks = 3) +
  scale_fill_manual(values = mags_colors_eco) +
  theme_classic() +
  theme(legend.position = "none",
        strip.placement = "outside",
        panel.border = element_blank()) +
  facet_grid(.~ecosystem, scales = "free") +
  coord_flip()
ggsave("plot_md_uSGBs_eco.pdf", plot_depth_usgbs, height = 8, width = 12)

@@

#match with the other ecosystems
mags_depth_kggbs_biom <- mags_metadata_bac[mags_metadata_bac$genus %in%
                                           mags_depth_kggbs_biom$genus, ] %>%
  dplyr::select(mag_name, mag_id, phylum, family, genus, ecosystem) %>%
  group_by(genus) %>%
  dplyr::mutate(tot_gen = n()) %>%
  dplyr::arrange(desc(tot_gen)) #nrow=535

#scale to 0-100%
mags_depth_kggbs_biom <- mags_depth_kggbs_biom %>%
  group_by(genus, ecosystem) %>%
  dplyr::mutate(sum = n()) %>% #sum genus per ecosystem
  dplyr::mutate(norm = sum / tot_gen * 100) %>%
  distinct(genus, ecosystem, .keep_all = TRUE) #reduce to uniques (plot scale!)

plot_depth_kggbs_biom <- ggplot(mags_depth_kggbs_biom,
                                aes(y = reorder(genus, norm), x = norm, fill = ecosystem)) +
  geom_col(position = position_stack(reverse = TRUE)) +
  scale_x_continuous(limits = c(0, 100), n.breaks = 5) +
  scale_fill_manual(values = mags_colors_eco) +
  labs(x = "% kGGBs", y = "Known genus from biomining") +
  theme_classic() +
  theme(legend.title = element_blank())
ggsave("plot_md_kGGBs_biom.pdf", plot_depth_kggbs_biom, height = 7, width = 6)

### --OTHER/PENDING TAXA-- ###
# Inspect unannotated family-level bins (uFGBs)? #1166
#mags_depth_ufgbs

# What about novel bacteria at genus level?
mags_depth_novel_gen <- mags_metadata_bac %>%
  dplyr::filter(genus == "") %>%
  group_by(phylum, genus, ecosystem) %>%
  dplyr::summarise(gen_unk = n()) %>%
  #arrange(desc(gen)) %>%
  ungroup() %>%
  group_by(genus) %>%
  dplyr::mutate(tot_gen = sum(gen_unk)) %>%
  dplyr::mutate(perc = gen_unk * 100 / tot_gen)

# Get unique classes of unknown genus by ecosystem
#Leptospirillia
#Sulfobacillia
mags_md_unk_gen2 <- mags_metadata_bac %>%
  dplyr::filter(genus == "") %>%
  group_by(class, ecosystem) %>%
  tally() %>%
  #slice(n()) %>%
  ungroup()
unique(mags_md_unk_gen2[mags_md_unk_gen2[, "ecosystem"] == "Biomining", "class"])

# Unique genus from biomining: Acidithrix, Ferrithrix, SCQM01
unique(mag_samples[mag_samples[, "family"] != "", "genus"])
mags_metadata_bac %>% filter(grepl("Acidithrix", genus, ignore.case = TRUE) == TRUE)

##plot
plot_md_unk_cla <- ggplot(mags_md_unk_gen2[mags_md_unk_gen2$n > 10, ],
                          aes(x = reorder(class, n), y = n, fill = ecosystem)) +
  geom_bar(stat = "identity") +
  labs(x = "Class", y = "Count") +
  scale_y_continuous(limits = c(0, 250)) +
  theme_minimal() +
  coord_flip() +
  facet_wrap(~ecosystem)
##
ggsave("plot_md_unk_class_eco.pdf", plot_md_unk_cla, height = 7, width = 10)

## Class of interest of acidic soils:
# Alphaproteobacteria, Gammaproteobacteria
# Acidobacteriae, Thermoleophilia, Sulfobacillia, Acidimicrobiia
table(mags_metadata_bac[mags_metadata_bac[, "phylum"] == "Proteobacteria", "class"])
table(mags_metadata_bac[mags_metadata_bac[, "class"] == "Acidimicrobiia", "order"])

## Family of interest reported for Systemix [ref 2022]:
# Pseudomonadaceae (mayor), Flavobacteriaceae, Erwiniaceae
mags_md_fams_acid <- mags_metadata_bac %>%
  filter(family == "Pseudomonadaceae") %>%
  group_by(ecosystem) %>%
  dplyr::summarise(count = n())
```

### QC (biomining)

```{r magsqc}
# Read QC file from CheckM
mags_qc <- read.csv("2-statistics/checkm_parsed_bins.tsv", sep = "\t")

@@

### Phylogenetic tree (biomining)

```{r magstree}
# Load GTDB taxonomy
gtdb_taxonomy <- read.csv("1-taxonomy/gtdbtk_classification.tsv", sep = "\t", stringsAsFactors = FALSE)

@@

fig_1 <- (fig_1a / fig_1b) +
  plot_annotation(tag_levels = list(c("a", "e", "b", "c", "d", "", "f"))) &
  theme(plot.tag =  element_text(face = "bold"))
ggsave("fig_1.pdf", fig_1, height = 14, width = 14)
rm(fig_1a, fig_1b)
#ggsave("fig_1.png", fig_1, dpi = 300, dev = "png", height = 14, width = 14, units = "cm")

##fig sup 1
fig_s1 <- { plot_depth_top_phyla | plot_depth_spec } +
  plot_layout(ncol = 2, widths = c(0.5, 1)) &
  theme(axis.title.x = element_text(margin = margin(t = -75, unit = "pt")))
fig_s1 <- { fig_s1 /
    { (plot_spacer() + plot_depth_usgbs + plot_spacer()) +
        plot_layout(ncol = 3, widths = c(0.5, 12, 2)) } #centered
  } + plot_annotation(tag_levels = list(c("a", "b", "c"))) &
  theme(plot.tag =  element_text(face = "bold"))
ggsave("fig_s1.pdf", fig_s1, height = 14, width = 14)
#ggsave("fig_s1.png", fig_s1, dpi = 300, dev = "png", height = 14, width = 14, units = "cm")
```

## Functional Annotation: KEGG KOs from KOfamScan

>>>>

---
title: "Systemix Article: Combined MAG-SMAG Profiling"
author: "Moises A. Rojas"
date: "2023-12-01 // mod. Mar 2024"
output: github_document
---

```{r setup}
#knitr::opts_knit$set(root.dir = '/mnt/beegfs/home/mrojas/DiGenomaLab/Systemix/analysis/mags-all/analysis/')
knitr::opts_knit$set(root.dir = '/Users/moises/Downloads/mags-all/')

suppressMessages(library(ape)) #tree basics
suppressMessages(library(treeio)) #join trees
suppressMessages(library(ggtree)) #tree plotting
suppressMessages(library(ggplot2)) #plots
suppressMessages(library(ggtreeExtra)) #geom_fruit
suppressMessages(library(ggnewscale)) #scales in ggplot2
suppressMessages(library(ggstar)) #geom_star for ggplot2
suppressMessages(library(tidyverse)) #strings, dataframes processing
suppressMessages(library(magrittr)) #pipes %>%
suppressMessages(library(tidytree)) #trees as dataframes
suppressMessages(library(RColorBrewer)) #color palettes
suppressWarnings(library(colorspace)) #colors manipulation
suppressWarnings(library(patchwork)) #side-by-side plotting
suppressWarnings(library(openxlsx)) #read excel
suppressWarnings(library(ComplexHeatmap)) #heatmap
suppressWarnings(library(PCAtools)) #pca
suppressWarnings(library(ggbiplot)) #pca plots
suppressWarnings(library(agricolae)) #fisher's test
suppressWarnings(library(ggpubr)) #stats & plots
suppressWarnings(library(dndscv)) #dN/dS of snps
suppressWarnings(library(clusterProfiler)) #gene enrichment
```

@@

```{r magdata}
# Read Systemix MAGs taxonomy and QC files
gtdb_taxonomy <- read.csv("1-taxonomy/gtdbtk_classification.tsv", sep = "\t", stringsAsFactors = FALSE)
mags_qc <- read.csv("2-statistics/checkm_parsed_bins.tsv", sep = "\t")

@@

# Parse columns
mag_samples <- dplyr::select(mag_samples, -c("taxonomy"))
mag_samples <- mag_samples %>%
  mutate_if(is.character, function(x) gsub("d__|p__|c__|o__|f__|g__|s__", "", x))
mag_samples <- mag_samples %>% separate(bin, "\\.", into = c("mag_name", "mag_id"))

# Define final attributes
mag_samples['mag_num'] <- nrow(mag_samples)
mag_samples['ecosystem'] <- "Biomining"
mag_samples['continent'] <- "South America"

# Load SMAG catalog (metadata from tables 1-2)
smag_md1 <- read.xlsx("/mnt/beegfs/home/mrojas/DiGenomaLab/Systemix/SMAG/Extended_Data_Table_1.xlsx")
smag_md2 <- read.xlsx("/mnt/beegfs/home/mrojas/DiGenomaLab/Systemix/SMAG/Extended_Data_Table_2.xlsx")

# Save attributes
smag_md1_sel <- smag_md1 %>%
  dplyr::select("mag_name", "mag_num", "Ecosystem", "Continent")
smag_md2_sel <- smag_md2 %>%
  dplyr::select("user_genome", "completeness", "contamination",
                "strain_heterogeneity", "length", "N50", "classification",
                "phylum", "class", "order", "family", "genus", "species")

@@

# Rename columns
colnames(smag_data)[6] <- "genome_size"
colnames(smag_data)[8] <- "kingdom"
colnames(smag_data)[16] <- "ecosystem"
colnames(smag_data)[17] <- "continent"

colnames(smag_data)[7] <- "kingdom"
colnames(smag_data)[14] <- "ecosystem"
colnames(smag_data)[15] <- "continent"

@@

# Save parsed SMAG metadata
if (file.exists("4-mags-smag/smag_data.tsv") == FALSE) {
  write.table(smag_data, "4-mags-smag/smag_data.tsv", sep = "\t", quote = FALSE, row.names = FALSE)
}

# Save biomining metadata (S15)
if (file.exists("4-mags-smag/mags_biomining.tsv") == FALSE) {
  write.table(mag_samples, "4-mags-smag/mags_biomining.tsv", sep = "\t", quote = FALSE, row.names = FALSE)
}

# Clean memory
rm(smag_md1, smag_md2, smag_md1_sel, smag_md2_sel)
#rm(list = ls(pattern = "^kegg_hm_cols_"))
```

## Section 1: In-depth taxonomy analysis

### Biomining MAGs vs SMAG taxa comparison

```{r magcombtaxa}
# Biomining provides a new phylum -> Nitrospirota_A
mag_samples_new_phy <- mag_samples[!(mag_samples$phylum %in% smag_data$phylum), ]

# Explore metadata: some bins are not assigned to ecosystems
mags_metadata %>%
  dplyr::filter(is.na(ecosystem)) %>%
  dplyr::group_by(mag_name) %>%
  dplyr::summarise() #n=401

# Keep only bacteria and defined ecosystems (discard archaea kingdom and non-NA eco.)
mags_metadata_bac <- mags_metadata %>%
  dplyr::filter(kingdom == "Bacteria" & !is.na(ecosystem)) #n=32975

# Save bacterial metadata to file
if (file.exists("4-mags-smag/mags_metadata_bac.tsv") == FALSE) {
  write.table(mags_metadata_bac, "4-mags-smag/mags_metadata_bac.tsv", sep = "\t", quote = FALSE, row.names = FALSE)
}

# Select top phyla and plot a histogram
mags_md_top_phy <- table(mags_metadata_bac$phylum) %>%
  sort(TRUE) %>%
  head(n = 40) %>%
  as.data.frame() %>%
  dplyr::rename(phy_top = Var1, freq = Freq)

plot_depth_top_phyla <- ggplot(mags_md_top_phy, aes(x = freq, y = reorder(phy_top, freq))) +
  geom_bar(stat = "identity", fill = "#459682") +
  geom_text(aes(label = freq), vjust = 0.5, hjust = -0.5, size = 2.5) +
  labs(x = "Number", y = "Phylum") +
  scale_x_continuous(limits = c(0, 10000)) +
  theme_classic()
ggsave("plot_md_top_phyla.pdf", plot_depth_top_phyla, height = 7, width = 6)

# Filter metadata..
# Set colors for ecosystem
#mags_colors_eco <- colorRampPalette(brewer.pal(12, "Paired"))(length(unique(mags_metadata_bac$ecosystem)))
#mags_colors_eco <- sample(mags_colors_eco)
mags_colors_eco <- c("#1F78B4", "#FF7F00", "#B2DF8A", "#E31A1C", "#A6CEE3",
                     "#6A3D9A", "#33A02C", "#CAB2D6", "#FDBF6F", "#FB9A99")

# Which bacteria from biomining sites are annotated as known?
mags_depth_spec_biom <- mags_metadata_bac %>%
  dplyr::select(mag_name, family, genus, species, ecosystem) %>%
  dplyr::filter(species != "" & ecosystem == "Biomining")

#are they present in other SMAG?
mags_depth_spec <- mags_metadata_bac[mags_metadata_bac$species %in%
                                     mags_depth_spec_biom$species, ] %>%
  dplyr::select(mag_name, mag_id, phylum, family, genus, species, ecosystem, continent) %>%
  dplyr::group_by(ecosystem)

plot_depth_spec <- ggplot(mags_depth_spec) +
  geom_bar(aes(x = species, fill = phylum)) +
  labs(x = "", y = "Number of kSGBs from biomining") +
  scale_fill_manual(values = "#78Acc2") +
  theme_classic() +
  theme(legend.title = element_blank(),
        legend.position = c(0.88, 0.12),
        strip.background = element_blank()) +#axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
  facet_grid(.~ecosystem) +
  coord_flip()
ggsave("plot_md_biom_species.pdf", plot_depth_spec, height = 2.5, width = 8)

# Which ecosystem has more known species (# and ratio)?
mags_depth_spec_eco <- mags_metadata_bac %>%
  dplyr::filter(species != "") %>%
  dplyr::group_by(species, ecosystem) %>%
  dplyr::summarise(count_sp = table(species)) %>% #.groups = "drop"
  ungroup() %>%
  dplyr::group_by(ecosystem) %>%
  dplyr::summarise(count_eco = sum(count_sp)) %>%
  dplyr::arrange(desc(count_eco)) #max 1821

@@

# What about unknown/known species-level genome bins (SGBs) at phylum level?
mags_depth_sgbs <- mags_metadata_bac %>%
  dplyr::group_by(phylum, species) %>%
  dplyr::summarise(n_spe = n()) %>%
  ungroup() %>%
  dplyr::group_by(phylum) %>%
  dplyr::mutate(t_phy = sum(n_spe)) %>% ###
  dplyr::filter(species == "") %>%
  dplyr::mutate(diff_k = t_phy - n_spe) %>%
  dplyr::mutate(perc_u = n_spe * 100 / t_phy) %>%
  dplyr::arrange(desc(t_phy)) %>%
  pivot_longer(cols = c(n_spe, diff_k), names_to = "cond", values_to = "tot")

@@

plot_depth_sgbs_p <- ggplot(mags_depth_sgbs_p[1:60, ],
                            aes(x = reorder(phylum, t_phy), y = tot_p, fill = cond_p)) +
  geom_col(position = position_stack(reverse = TRUE)) +
  scale_fill_manual(name = "", labels = c("kSGB", "uSGB"), values = c("#DD7E8B", "#459682")) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(x = "", y = "% of species") +
  theme_classic() +
  theme(legend.position = "none",
        axis.text.y = element_blank(),
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank()) +
  coord_flip()
#ggsave("plot_md_SGBs_phylum_p.pdf", plot_depth_sgbs_p, height = 7, width = 5)

@@

# And unknown/known SGBs for biomining?
mags_depth_sgbs_biom <- mags_metadata_bac %>%
  dplyr::filter(ecosystem == "Biomining") %>%
  dplyr::group_by(phylum, species) %>%
  dplyr::summarise(n_spe = n()) %>%
  ungroup() %>%
  dplyr::group_by(phylum) %>%
  dplyr::mutate(t_phy = sum(n_spe)) %>%
  dplyr::filter(species == "") %>%
  dplyr::mutate(diff_k = t_phy - n_spe) %>%
  dplyr::mutate(perc_u = n_spe * 100 / t_phy) %>%
  dplyr::arrange(desc(t_phy)) %>%
  pivot_longer(cols = c(n_spe, diff_k), names_to = "cond", values_to = "tot")

@@

# What is the distribution of uSGBs by ecosystem?
mags_depth_usgbs_eco <- mags_metadata_bac %>%
  dplyr::filter(species == "") %>%
  dplyr::group_by(phylum, ecosystem) %>%
  dplyr::summarise(num = table(ecosystem)) %>%
  dplyr::mutate(phylum = replace(phylum, num < 5, "Other")) #group lows as "other" -> summarise?

plot_depth_usgbs <- ggplot(mags_depth_usgbs_eco,
                           aes(x = num, y = reorder(phylum, num), fill = ecosystem)) +
  geom_col() +
  #geom_text(aes(label = num), vjust = 0.5, hjust = -0.5, size = 2.5) +
  labs(x = "Number of uSGBs", y = "Phylum") +
  scale_x_continuous(n.breaks = 3) +
  scale_fill_manual(values = mags_colors_eco) +
  theme_classic() +
  theme(legend.position = "none",
        strip.placement = "outside",
        strip.background = element_blank(),
        panel.border = element_blank()) +
  facet_grid(.~ecosystem, scales = "free")
ggsave("plot_md_uSGBs_eco.pdf", plot_depth_usgbs, height = 8, width = 12)

@@

#match with the other ecosystems
mags_depth_kggbs_biom <- mags_metadata_bac[mags_metadata_bac$genus %in%
                                           mags_depth_kggbs_biom$genus, ] %>%
  dplyr::select(mag_name, mag_id, phylum, family, genus, ecosystem) %>%
  dplyr::group_by(genus) %>%
  dplyr::mutate(tot_gen = n()) %>%
  dplyr::arrange(desc(tot_gen)) #nrow=535

#scale to 0-100%
mags_depth_kggbs_biom <- mags_depth_kggbs_biom %>%
  dplyr::group_by(genus, ecosystem) %>%
  dplyr::mutate(sum = n()) %>% #sum genus per ecosystem
  dplyr::mutate(norm = sum / tot_gen * 100) %>%
  distinct(genus, ecosystem, .keep_all = TRUE) #reduce to uniques (plot scale!)

plot_depth_kggbs_biom <- ggplot(mags_depth_kggbs_biom,
                                aes(y = reorder(genus, norm), x = norm, fill = ecosystem)) +
  geom_col(position = position_stack(reverse = TRUE)) +
  scale_x_continuous(limits = c(0, 101), n.breaks = 5) + #check!
  scale_fill_manual(values = mags_colors_eco) +
  labs(x = "% kGGBs", y = "Known genus from biomining") +
  theme_classic() +
  theme(legend.title = element_blank())
ggsave("plot_md_kGGBs_biom.pdf", plot_depth_kggbs_biom, height = 7, width = 6)

### --EXPLORE OTHER TAXA-- ###
#inspect unannotated family-level bins (uFGBs)? #1166
#what about novel bacteria at genus level?
mags_depth_novel_gen <- mags_metadata_bac %>%
  dplyr::filter(genus == "") %>%
  dplyr::group_by(phylum, genus, ecosystem) %>%
  dplyr::summarise(gen_unk = n()) %>%
  #dplyr::arrange(desc(gen)) %>%
  ungroup() %>%
  dplyr::group_by(genus) %>%
  dplyr::mutate(tot_gen = sum(gen_unk)) %>%
  dplyr::mutate(perc = gen_unk * 100 / tot_gen)

#get unique classes of unknown genus by ecosystem
#Leptospirillia, Sulfobacillia
mags_md_unk_gen2 <- mags_metadata_bac %>%
  dplyr::filter(genus == "") %>%
  dplyr::group_by(class, ecosystem) %>%
  tally() %>%
  #slice(n()) %>%
  ungroup()
unique(mags_md_unk_gen2[mags_md_unk_gen2[, "ecosystem"] == "Biomining", "class"])

#unique genus from biomining: Acidithrix, Ferrithrix, SCQM01
unique(mag_samples[mag_samples[, "family"] != "", "genus"])
mags_metadata_bac %>% filter(grepl("Acidithrix", genus, ignore.case = TRUE) == TRUE)

plot_md_unk_cla <- ggplot(mags_md_unk_gen2[mags_md_unk_gen2$n > 10, ],
                          aes(x = reorder(class, n), y = n, fill = ecosystem)) +
  geom_bar(stat = "identity") +
  labs(x = "Class", y = "Count") +
  scale_y_continuous(limits = c(0, 250)) +
  theme_minimal() +
  coord_flip() +
  facet_wrap(~ecosystem)
ggsave("plot_md_unk_class_eco.pdf", plot_md_unk_cla, height = 7, width = 10)

#class of interest for acidic soils:
#Alphaproteobacteria, Gammaproteobacteria
#Acidobacteriae, Thermoleophilia, Sulfobacillia, Acidimicrobiia
table(mags_metadata_bac[mags_metadata_bac[, "phylum"] == "Proteobacteria", "class"])
table(mags_metadata_bac[mags_metadata_bac[, "class"] == "Acidimicrobiia", "order"])

#family of interest reported for Systemix [ref 2022]:
#Pseudomonadaceae (mayor), Flavobacteriaceae, Erwiniaceae
mags_md_fams_acid <- mags_metadata_bac %>%
  dplyr::filter(family == "Pseudomonadaceae") %>%
  dplyr::group_by(ecosystem) %>%
  dplyr::summarise(count = n())
```

### QC

```{r magbiomqc}
# Read QC file from CheckM
mags_qc <- read.csv("2-statistics/checkm_parsed_bins.tsv", sep = "\t")

@@

### Phylogenetic tree

```{r magbiomtree}
# Load GTDB taxonomy
gtdb_taxonomy <- read.csv("1-taxonomy/gtdbtk_classification.tsv", sep = "\t", stringsAsFactors = FALSE)

@@

fig_1 <- (fig_1a / fig_1b) +
  plot_annotation(tag_levels = list(c("a", "e", "b", "c", "d", "", "f"))) &
  theme(plot.tag =  element_text(face = "bold"))
ggsave("fig_1.pdf", fig_1, height = 14, width = 14)
rm(fig_1a, fig_1b)

#fig sup 1
fig_s1 <- { plot_depth_top_phyla |
    (plot_depth_usgbs / plot_depth_spec + plot_layout(heights = c(0.85, 0.15))) } +
  plot_layout(ncol = 2, widths = c(0.25, 1)) +
  plot_annotation(tag_levels = list(c("a", "b", "c"))) &
  theme(plot.tag =  element_text(face = "bold"))
ggsave("fig_s1.pdf", fig_s1, height = 8, width = 16)
#ggsave("fig_s1.png", fig_s1, height = 8, width = 16)
```

## Section 2: Functional Annotation

### Pre-process matrix of KEGG KOs from KOfamScan

```{r processkos}
# Read output matrix from KOfamScan
kegg_annot <- read.csv("/mnt/beegfs/home/mrojas/DiGenomaLab/Systemix/analysis/kegg/merge_hits/kofamscan_hits.tsv", sep = "\t", check.names = FALSE)
colnames(kegg_annot)[2] <- "KO.definition"

# Extract KO definitions and set KOs as row names
kegg_kos <- data.frame(KO = kegg_annot$KO, Definition = kegg_annot$KO.definition)
kegg_annot <- kegg_annot %>%
  dplyr::select(-c("KO.definition")) %>%
  data.frame(row.names = 1, check.names = FALSE)

# Set definitions for pathways of interest
kegg_paths <- kegg_kos %>%
  #sulf: to keep sulfur, sulfate, sulfane, sulfide, sulfo_
  #soxr: SoxR 2Fe-2S cluster
  #nitrogen: a negative marker (ubiquous; ex: MerR family)
  dplyr::filter(grepl("iron|sulf|soxr|copper|nitrogen",
                      Definition, ignore.case = TRUE) == TRUE) %>%
  dplyr::select(KO, Definition)
